### 2025-07-18 paper/quarto_aaai/paper.tex
Finish editorial pass

This section introduces the counterfactual training framework. As
mentioned above, CT applies ideas from contrastive and robust learning
to counterfactual explanations in order to produce models whose learned
representations align with plausible explanations that further comply
with user-defined actionability constraints.

@@

Our main results for plausibility and actionability for \emph{MLP}
models are summarised in Table~\ref{tbl-main}, which presents
counterfactual outcomes grouped by dataset along with standard errors
averaged across bootstrap samples. Asterisks (\(^*\)) indicate that the
observed effects are statistically significant: bootstrapped
99\%-confidence intervals of differences in mean outcomes do \emph{not}
include zero.

The first two columns (\(\text{IP}\) and \(\text{IP}^*\)) show the
percentage reduction in implausibility for our two metrics when using CT
on top of the weak baseline. As an example, consider the first row
(\emph{LS}): the observed positive values indicate that faithful
counterfactuals are around 30-55\% more plausible for models trained
with CT, in line with our observations in Figure~\ref{fig-poc} in panel
(b) compared to panel (a).

@@

\emph{CT tends to improve actionability in the presence of immutable
features, but the exact effects can be challenging to establish due to
feature interactions.}

For synthetic datasets, we always protect the first feature; for all
real-world tabular datasets we could identify and protect an \emph{age}
variable; for \emph{MNIST}, we protect the five upper and lower pixel
rows of the full image. Reductions in costs are overwhelmingly positive
and significant of up to around 60\% for \emph{GMSC}. While the
estimated cost increase for \emph{Adult} and \emph{MNIST}, these effects
are not significant and Figure~\ref{fig-mnist} (columns 6-10)
demonstrates that in the latter case, CT does have the expected effect:
sensitivity to protected features as per integrated gradients is
drastically reduced; details of this experiment are reported in the
supplementary appendix. In the case of \emph{Cred}, average costs
increase, likely because any potential benefits from protecting the
\emph{age} are outweighed by the increase in costs required for greater
plausibility.

@@

State-of-the-art machine learning models are prone to learning complex
representations that cannot be interpreted by humans. Existing
explainability solutions cannot guarantee that explanations agree with
these learned representation. As a step towards addressing this
challenge, we introduce counterfactual training, a novel training regime
that incentivizes highly-explainable models. Through extensive
experiments, we demonstrate that CT satisfies this objective while
promoting robustness and preserving the predictive performance of
models. Explanations generated from CT-based models are both more
plausible (compliant with the underlying data-generating process) and
more actionable (compliant with user-specified mutability constraints),
and thus meaningful to their recipients. We also show that CT can be
used to fine-tune conventionally-trained models and achieve similar
gains. Lastly, our work highlights the value of simultaneously improving
models and their explanations.

>>>>

This section introduces the counterfactual training framework, applying
ideas from contrastive and robust learning to counterfactual
explanations. CT produces models whose learned representations align
with plausible explanations that comply with user-defined actionability
constraints.

@@

Our main results for plausibility and actionability for \emph{MLP}
models are summarised in Table~\ref{tbl-main} that presents
counterfactual outcomes grouped by dataset along with standard errors
averaged across bootstrap samples. Asterisks (\(^*\)) are used when the
bootstrapped 99\%-confidence interval of differences in mean outcomes
does \emph{not} include zero, so the observed effects are statistically
significant at the 0.01 level.

The first two columns (\(\text{IP}\) and \(\text{IP}^*\)) show the
percentage reduction in implausibility for our two metrics when using CT
on top of the weak baseline. As an example, consider the first row for
\emph{LS} data: the observed positive values indicate that faithful
counterfactuals are around 30-55\% more plausible for models trained
with CT, in line with our observations in Figure~\ref{fig-poc} in panel
(b) compared to panel (a).

@@

\emph{CT tends to improve actionability in the presence of immutable
features, but the exact effect is sometimes hard to establish due to,
e.g., feature interactions.}

For synthetic datasets, we always protect the first feature; for all
real-world tabular datasets we could identify and protect an \emph{age}
variable; for \emph{MNIST}, we protect the five upper and lower pixel
rows of the full image. Reductions in costs are overwhelmingly positive
and significant, of up to around 60\% for \emph{GMSC}. While the
estimated cost increase for \emph{Adult} and \emph{MNIST}, these effects
are not significant and Figure~\ref{fig-mnist} (columns 6-10)
demonstrates that in the latter case, CT does have the expected effect:
sensitivity to protected features as per integrated gradients is
drastically reduced; details of this experiment are reported in the
supplementary appendix. In the case of \emph{Cred}, average costs
increase, likely because any potential benefits from protecting the
\emph{age} are outweighed by the increase in costs required for greater
plausibility.

@@

State-of-the-art machine learning models are prone to learning complex
representations that cannot be interpreted by humans. Existing
explainability solutions cannot guarantee that explanations agree with
these learned representation. As a step towards addressing this
challenge, we introduce counterfactual training, a novel training regime
that integrates recent advances in contrastive learning, adversarial
robustness, and counterfactual explanations to incentivize
highly-explainable models. Through extensive experiments, we demonstrate
that CT satisfies this goal while preserving the predictive performance
and promoting robustness of models. Explanations generated from CT-based
models are both more plausible (compliant with the underlying
data-generating process) and more actionable (compliant with
user-specified mutability constraints), and thus meaningful to their
recipients. In turn, our work highlights the value of simultaneously
improving models and their explanations.

### 2025-07-18 paper/sections/experiments.qmd
Finish editorial pass

Our main results for plausibility and actionability for *MLP* models are summarised in @tbl-main, which presents counterfactual outcomes grouped by dataset along with standard errors averaged across bootstrap samples. Asterisks ($^*$) indicate that the observed effects are statistically significant: bootstrapped 99%-confidence intervals of differences in mean outcomes do *not* include zero.

The first two columns ($\text{IP}$ and $\text{IP}^*$) show the percentage reduction in implausibility for our two metrics when using CT on top of the weak baseline. As an example, consider the first row (*LS*): the observed positive values indicate that faithful counterfactuals are around 30-55% more plausible for models trained with CT, in line with our observations in @fig-poc in panel (b) compared to panel (a).

@@

### Actionability (RQ2). {#sec-act}
*CT tends to improve actionability in the presence of immutable features, but the exact effects can be challenging to establish due to feature interactions.*

For synthetic datasets, we always protect the first feature; for all real-world tabular datasets we could identify and protect an *age* variable; for *MNIST*, we protect the five upper and lower pixel rows of the full image. Reductions in costs are overwhelmingly positive and significant of up to around 60% for *GMSC*. While the estimated cost increase for *Adult* and *MNIST*, these effects are not significant and @fig-mnist (columns 6-10) demonstrates that in the latter case, CT does have the expected effect: sensitivity to protected features as per integrated gradients is drastically reduced; details of this experiment are reported in the supplementary appendix. In the case of *Cred*, average costs increase, likely because any potential benefits from protecting the *age* are outweighed by the increase in costs required for greater plausibility.

>>>>

Our main results for plausibility and actionability for *MLP* models are summarised in @tbl-main that presents counterfactual outcomes grouped by dataset along with standard errors averaged across bootstrap samples. Asterisks ($^*$) are used when the bootstrapped 99%-confidence interval of differences in mean outcomes does *not* include zero, so the observed effects are statistically significant at the 0.01 level.

The first two columns ($\text{IP}$ and $\text{IP}^*$) show the percentage reduction in implausibility for our two metrics when using CT on top of the weak baseline. As an example, consider the first row for *LS* data: the observed positive values indicate that faithful counterfactuals are around 30-55% more plausible for models trained with CT, in line with our observations in @fig-poc in panel (b) compared to panel (a).

@@

### Actionability (RQ2). {#sec-act}
*CT tends to improve actionability in the presence of immutable features, but the exact effect is sometimes hard to establish due to, e.g., feature interactions.*

For synthetic datasets, we always protect the first feature; for all real-world tabular datasets we could identify and protect an *age* variable; for *MNIST*, we protect the five upper and lower pixel rows of the full image. Reductions in costs are overwhelmingly positive and significant, of up to around 60% for *GMSC*. While the estimated cost increase for *Adult* and *MNIST*, these effects are not significant and @fig-mnist (columns 6-10) demonstrates that in the latter case, CT does have the expected effect: sensitivity to protected features as per integrated gradients is drastically reduced; details of this experiment are reported in the supplementary appendix. In the case of *Cred*, average costs increase, likely because any potential benefits from protecting the *age* are outweighed by the increase in costs required for greater plausibility.
